---
title: "Modular Matrix Operations of Mutational Events"
author: 
    - name: Robersy Sanchez
      affiliation: Department of Biology.
                   Pennsylvania State University, University Park, PA 16802
      email: rus547@psu.edu
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 4
        toc_float:
                collapsed: false
                smooth_scroll: true
        number_sections: true
        theme: united 
        geometry: margin=0.8in
        highlight: tango
toc_float: true
toc_depth: 4
abstract: |
    A fast introduction into the analysis of DNA mutational events 
    as modular matrix operation, i.e., as automorphisms over a ring.
vignette: >
  %\VignetteIndexEntry{Get started-with GenomAutomorphism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


```{r set-options, echo=FALSE, cache=FALSE}
options(width = 120)
```

# Introduction

In the Abelian p-group defined on $\mathbb{Z}_{p^{\alpha_{i}}_i}$  the 
endomorphisms $\eta_i \in End(\mathbb{Z}_{p^{\alpha_{i}}_i})$ are described as
functions $f(x)=kx$ mod $p^{\alpha_{i}}_i$, where $k$ and $x$ are elements 
from the set of integers modulo $p^{\alpha_{i}}_i$.

Next, we proceed to load the R package required libraries
```{r lib}
library(GenomAutomorphism)
```

# Modular operation on homocyclic group

The mutational events transforming the codon sequence ATACCCATGGCCAAC into the
sequence ACACCCATGACCAAC can be represented by means of automorphisms.


In the cube ACGT the codon sequence ATACCCATGGCCAAC is represented by the 
vector $(48,21,50,25,1) \in (\mathbb{Z}_{2^6})^5$, where $(\mathbb{Z}_{2^6})^5$
stands for the homocyclic Abelian group of vectors of integers modulo 
$\mathbb{Z}_{2^6}$, while the sequence ACACCCATGACCAAC is represented by
$(16,21,50,17,1)$.

To translate the DNA sequence into numerical vectors from 
$(\mathbb{Z}_{2^6})^5$ they are previously incorporated into a 
[DNAStringSet](https://rdrr.io/bioc/Biostrings/man/XStringSet-class.html) 
object.

```{r dna}
dna <- DNAStringSet(c('ATACCCATGGCCAAC', 'ACACCCATGACCAAC'))
dna
```


```{r mod}
cd <- codon_coord(dna, cube = "ACGT", group = "Z64")
cd
```

A list of coordinates is retrieved from object *cd*
```{r cd1}
cd <- coordList(get_coord(cd))
cd
```

## Homomorphism matrices

Most of the mutational events in small DNA segments can be represented by 
means homomorphisms with which, in turns, are presented by diagonal matrices.
The diagonal elements of matrix a transforming coordinate elements from
*coord1* into *coord2* are estimated as shown below
```{r mle}
dg <- modlineq(cd$coord1, cd$coord2, 64)
dg
```

That is, we have the matrix:
```{r m}
diag(dg)
```
The multiplication of vector *coord1* by the matrix *diag(dg)* yields vector
coord2:
```{r m1}
(cd$coord1 %*% diag(dg)) %% 64
# Or
cat("\n---- \n")
(cd$coord1 %*% diag(dg)) %% 64 == cd$coord2

```

```{r m2}
dna2 <- DNAStringSet(c('GACAGAGCAGTATTAGCTTCACAC', 'GAAAAC---GTATTA---TCAAAG'))
dna2
```

```{r mod2}
cd2 <- codon_coord(dna2, cube = "ACGT", group = "Z64")
cd2
```

```{r cd}
cd2$coord2[ is.na(cd2$coord2) ] <- 0
cd2
```

```{r cd2}
cd2 <- coordList(get_coord(cd2))
cd2
```


Coordinates of *seq2* can be obtained from *seq1* solving the equation:
$seq1 \cdot x = seq2$ mod $n$
```{r m3}
modulo <- c(64,125,64,64,64,64,64,64)
coord1 <- cd2$coord1
coord2 <- cd2$coord2

dg <- modlineq(coord1, coord2, modulo)
dg
```

Or in a form of diagonal matrix:
```{r d}
m  <- diag(dg)
```

That is, the matrix multiplication below yields coordinates of *seq2*.
```{r m4}
(coord1 %*% m) %% modulo
```

Or
```{r m5}
(coord1 %*% m) %% modulo == coord2
```

## Affine Transformations

Next, we search for the homomorphims transforming coordinates coord2 into
coodinates coord1:
```{r m6}
modulo <- c(64,125,64,64,64,64,64,64)

dg1 <- modlineq(a = coord2, b = coord1, n = modulo, no.sol = 0)
dg1
```

This is an affine transformation, which involves the sum of a homomorphism:
```{r hom}
m1 <- diag(dg1$diag)
m1
```

Plus a translation. That is:
```{r cheq}
(coord2 %*% m1 + dg1$translation) %% modulo

# Or
cat("\n---- \n")

(coord2 %*% m1 + dg1$translation) %% modulo == coord1
```

However, the solution is not unique. For example, setting no.sol = 1, we have
another solution
```{r m7}
dg2 <- modlineq(a = coord2, b = coord1, n = modulo, no.sol = 1)
dg2
```

```{r cheq1}
m2 <- diag(dg2$diag)

(coord2 %*% m2 + dg2$translation) %% modulo

# Or
cat("\n---- \n")

(coord2 %*% m2 + dg2$translation) %% modulo == coord1
```

Since the second element of *seq2* has coordinate is 1, to build a homomorphism that 
transform coordinates from *seq2* into coordinates from *seq1*, we can just sum 
the translation vector to the second column of matrix m2:

```{r m8}
m3 <- m2
m3[, 2] <- (m2[, 2] + dg2$translation) %% modulo
m3
```

It is easy to see that:
```{r m9}
(coord2 %*% t(m3)) %% modulo
```

Or, as a column vector:
```{r m10}
(m3 %*% coord2) %% modulo
```




